---
title: 'Initial Data Inspection: Prosthetic Infection Immune Profiling'
author: "Bruno Álvarez"
date: "2026-05-12"
output:
  pdf_document: 
    toc: true
    number_sections: true
    latex_engine: xelatex
  geometry: margin=1n
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)

library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(janitor)
library(ggplot2)
library(skimr)
library(naniar)
library(knitr)
library(kableExtra)
library(here)
library(writexl)
```

The aim of this first report is to do an initial inspection and a sanity check to fix errors on the data set to ensure it is usable for a proper analysis. THe analysis will be performed on the clinical databases and the FlowSOM output database

Main objectives to verify:
- Available Excel sheets
- Structure of the FlowSOM metacluster frequency table
- Structure of the metacluster marker profile table
- Structure of the clinical database
- Look for the sample identifiers
- Initial missigness patterns
- Feasibility of building a merged clinical-immune master table


Storing the file paths
```{r file-paths}
flowsom_file <- here("Data", "Trauma_FlowSOM_Report.xlsx")
clinical_file <- here("Data", "Estuido_Inmunidad_DB.xlsx")

flowsom_file
clinical_file
```

Overviewing datasets
```{r excel-sheet-overview}
flowsom_sheets <- excel_sheets(flowsom_file)
clinical_sheets <- excel_sheets(clinical_file)

flowsom_sheets
clinical_sheets
```

Loading the datasets
```{r dataset-loading}
clinical_demografics <- read_excel(clinical_file, sheet = "DemográficosClínicos", skip = 1)
clinical_bioq <- read_excel(clinical_file, sheet = "Bioquímicas", skip = 1)
clinical_immune <- read_excel(clinical_file, sheet = "Principales (Inmunidad)")
```

Cleaning each dataset columns
```{r dataset-column-cleaning}
clinical_demo_clean <- clinical_demografics %>% clean_names()
clinical_bioq_clean <- clinical_bioq %>% clean_names()
clinical_immune_clean <- clinical_immune %>% clean_names()
```

Proceeding to inspect the different clinical data sets, however I stop paying attention to the clinical Immune one because its just the FlowSOM data set wrongly organized and stored

Inspecting dimensions
```{r inspecting-the-dimensions-on-the-clinical-datasets}
dim(clinical_demo_clean)
dim(clinical_bioq_clean)
dim(clinical_immune_clean)
```

checking the column names
```{r checking-colnames}
print("- DEMOGRAFICS COLUMN NAMES")
colnames(clinical_demo_clean)
print("- BIOQUIMICA COLUMN NAMES")
colnames(clinical_bioq_clean)
```

taking a look at the first rows as examples
```{r cheacking-first-rows}
head(clinical_demo_clean)
head(clinical_bioq_clean)
```

The patient ID column is called "codigo_interno" so first I will double check it

```{r checking-patient-id}
"codigo_interno" %in% colnames(clinical_demo_clean)
"codigo_interno" %in% colnames(clinical_bioq_clean)
```

Taking a look at some of the IDs
```{r head-check}
clinical_demo_clean %>%
  select(codigo_interno) %>%
  head(20)

clinical_bioq_clean %>%
  select(codigo_interno) %>%
  head(20)

```

Checking all the rows to look for repeated IDs or rows with missing ID
```{r repeated-and-missing-id-sanity-check}
clinical_demo_clean %>%
  summarise(
    rows = n(),
    unique_ids = n_distinct(codigo_interno),
    missing_ids = sum(is.na(codigo_interno))
  )

clinical_bioq_clean %>%
  summarise(
    rows = n(),
    unique_ids = n_distinct(codigo_interno),
    missing_ids = sum(is.na(codigo_interno))
  )
```

Double checking for duplicates in the IDs
```{r doublle-check-for-duplicates}
clinical_demo_clean %>%
  count(codigo_interno) %>%
  filter(n > 1)

clinical_bioq_clean %>%
  count(codigo_interno) %>%
  filter(n > 1)
```

Checking if all patients in demografics are in bioquimica

this will show patients in DemograficosClínicos but not in Bioquímicas
```{r checking-if-patients-match}
clinical_demo_clean %>%
  anti_join(clinical_bioq_clean, by = "codigo_interno")
```
This will show patients in "Bioquímicas" but not in "DemograficosClínicos"
```{r}
clinical_bioq_clean %>%
  anti_join(clinical_demo_clean, by = "codigo_interno")
```

Since in both cases there is no table generated it means they 100% match

Taking a look now at the types of variables
```{r checking-variable-types}
str(clinical_demo_clean)
str(clinical_bioq_clean)
```

```{r checking-variable-types-clean}
glimpse(clinical_demo_clean)
glimpse(clinical_bioq_clean)
```

We can see that the variable "numero_de_intervenciones" is suposed to be just numbers but its a "chr" type variable instead of a "dbl" type variable, this is because in some cases not only the number is listed but also the type of interventions. Therefore I thought of the solution to preserve that column but also generate two derivates one containing just the number and anotherone containing the types of interventions.

```{r interventions-reframed}
clinical_demo_clean <- clinical_demo_clean %>%
  mutate(
    numero_de_intervenciones_original = numero_de_intervenciones,
    
    numero_intervenciones_n = str_extract(
      numero_de_intervenciones,
      "^\\d+(\\.\\d+)?"
    ) %>%
      as.numeric(),
    
    detalle_intervenciones = str_extract(
      numero_de_intervenciones,
      "\\(.*\\)"
    ) %>%
      str_remove_all("^\\(|\\)$") %>%
      str_squish()
  )
```

vhecking if the reframed worked
```{r checking}
clinical_demo_clean %>%
  select(
    codigo_interno,
    numero_de_intervenciones_original,
    numero_intervenciones_n,
    detalle_intervenciones
  ) %>%
  head(20)
```

Due to the format of the excel there was a column name lost, the column "x10" should be called "Variables principales (inmunofenotipo)"
```{r renaming column}
clinical_demo_clean <- clinical_demo_clean %>% rename(Variables_principales_inmunofenotipo = x10)
```

checking if it worked
```{r checking column}
glimpse(clinical_demo_clean)
```

Now that the clinical data has passed a sanity check we can create a master clinical data
```{r joining-clinical-datasets}
clinical_master <- clinical_demo_clean %>%
  full_join(clinical_bioq_clean, by = "codigo_interno")
```

checking how the joined dataset looks
```{r checking-the-joined-dataset}
dim(clinical_master)

clinical_master %>%
  select(codigo_interno) %>%
  summarise(
    rows = n(),
    unique_ids = n_distinct(codigo_interno),
    missing_ids = sum(is.na(codigo_interno))
  )
```

Now lets save the dataset into a CSV file
```{r saving-master-clinical-dataset}
dir.create(here("data", "processed"), showWarnings = FALSE, recursive = TRUE)

write.csv(
  clinical_master,
  here("data", "processed", "clinical_master_preliminary.csv"),
  row.names = FALSE
)
```


Proceding to import the FlowSOM database, with a database for each of its sheets ("MC Frequencies (%)" "MC Marker Profiles" "Methods"). However, I exclude the "Methods" database since its just a docs file emptied into an excel file
```{r importing FlowSOM}
MC_frequencies <- read_excel(flowsom_file, sheet = "MC Frequencies (%)", skip = 0)
MC_markers <- read_excel(flowsom_file, sheet = "MC Marker Profiles", skip = 0)
```

Cleaning the column names
```{r clean flowsom dataset}
MC_frequencies_clean <- MC_frequencies %>% clean_names()
MC_markers_clean <- MC_markers %>% clean_names()
```

Inspecting dimensions
```{r dim}
dim(MC_frequencies_clean)
dim(MC_markers_clean)
```

```{r}
colnames(MC_frequencies_clean)
colnames(MC_markers_clean)
```

```{r}
head(MC_frequencies_clean)
head(MC_markers_clean)
```

Sample "71" is actually the sample "1" so its name should be changed from "P71" to "P1" in MC_frequencies_clean dataset
```{r renaming sample 71}
MC_frequencies_clean <- MC_frequencies_clean %>%
  mutate(patient = if_else(patient == "P71", "P1", patient))
```

lets check if it changed successfully
```{r checking changed id}
print(MC_frequencies_clean$patient)

a <- FALSE
for (x in MC_frequencies_clean$patient) {
  if (x == "P71"){
    print("changed unsuccessfully")
  }
  else {
    a <- TRUE
  }
}
```

Since the column joining datasets is named "codigo_interno" we change the column name "patient" in "MC_frequencies_clean" to "codigo_interno"
```{r renaming id column}
MC_frequencies_clean <- MC_frequencies_clean %>% rename(codigo_interno = patient)
```

CHecking if it changed properly
```{r column change check}
glimpse(MC_frequencies_clean)
```

With the MC_frequencies data set checked we proceed to join all the data together, leaving it ready for a primary analysis analysis
```{r}
Trauma_dataset <- clinical_master %>%
  full_join(MC_frequencies_clean, by = "codigo_interno")
```


Checking the joined data set
```{r}
dim(Trauma_dataset)

Trauma_dataset %>%
  select(codigo_interno) %>%
  summarise(
    rows = n(),
    unique_ids = n_distinct(codigo_interno),
    missing_ids = sum(is.na(codigo_interno))
  )
```
There is a patient from which there was no sample, consequently it has no immunological data, due to that I will preserve it in the Trauma data set but will create a new data set excluding it for the analysis
```{r}
Trauma_analysis_dataset <- Trauma_dataset %>% slice(-5)
```

Checking if it worked
```{r}
dim(Trauma_dataset)
dim(Trauma_analysis_dataset)
```

Storing both data sets into a csv file
```{r}
write.csv(
  Trauma_dataset,
  here("Data", "Trauma_dataset.csv"),
  row.names = FALSE
)

write.csv(
  Trauma_analysis_dataset,
  here("Data", "Trauma_analysis_dataset.csv"),
  row.names = FALSE
)

write.csv(
  MC_markers_clean,
  here("Data", "MC_markers.csv"),
  row.names = FALSE
)
```