---
title: 'Quaity Control: Prosthetic Infection Immune Profiling'
author: "Bruno Álvarez"
date: "2026-05-14"
output:
  pdf_document: 
    toc: true
    number_sections: true
    latex_engine: xelatex
  geometry: margin=1n
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)

library(tidyverse)
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(janitor)
library(ggplot2)
library(skimr)
library(naniar)
library(knitr)
library(kableExtra)
library(here)
library(writexl)
library(pheatmap)
library(ggrepel)
library(broom)
```

Quality control of the data to clean it and leave it prepared

The first step is to map the objects, confirm provenance, identify true FlowSOM frequency columns, then QC missingness and compositional integrity.

Lets start by loading the datasets
```{r}
Trauma_data <- read.csv("/home/bruno/Academia/Master/TFM/proyecto_Trauma/resultados/Data/Trauma_analysis_dataset.csv")
MC_markers <- read.csv("/home/bruno/Academia/Master/TFM/proyecto_Trauma/resultados/Data/MC_markers.csv")
```

Lets take a look at them now
```{r}
dim(Trauma_data)
colnames(Trauma_data)
glimpse(Trauma_data)
```

```{r}
dim(MC_markers)
colnames(MC_markers)
glimpse(MC_markers)
```

load the frequencies and markers from the excel to backtrack check
```{r}
excel_frequencies <- read_excel(
  here("Data", "Trauma_FlowSOM_Report.xlsx"),
  sheet = "MC Frequencies (%)"
) %>%
  clean_names() %>%
  mutate(patient = if_else(patient == "P71", "P1", patient)) %>%
  rename(codigo_interno = patient)

excel_markers <- read_excel(
  here("Data", "Trauma_FlowSOM_Report.xlsx"),
  sheet = "MC Marker Profiles"
) %>%
  clean_names()
```

before analysis the main issue is the duplicated columns:
nombre.x
nombre.y
nhc.x
nhc.y
vsg_mm_h.x
vsg_mm_h.y
crp_serum_mg_d_l.x
crp_serum_mg_d_l.y
d_dimero_plasma_mcg_l_feu.x
d_dimero_plasma_mcg_l_feu.y

Lets compare the duplicate columns
```{r}
compare_duplicate_columns <- function(df, x_col, y_col) {
  tibble(
    codigo_interno = df$codigo_interno,
    x = df[[x_col]],
    y = df[[y_col]],
    both_missing = is.na(x) & is.na(y),
    same_value = x == y | both_missing
  ) %>%
    filter(!same_value | is.na(same_value))
}

compare_duplicate_columns(Trauma_data, "nombre.x", "nombre.y")
compare_duplicate_columns(Trauma_data, "nhc.x", "nhc.y")
compare_duplicate_columns(Trauma_data, "vsg_mm_h.x", "vsg_mm_h.y")
compare_duplicate_columns(Trauma_data, "crp_serum_mg_d_l.x", "crp_serum_mg_d_l.y")
compare_duplicate_columns(Trauma_data, "d_dimero_plasma_mcg_l_feu.x", "d_dimero_plasma_mcg_l_feu.y")
```

Now just keep the existing value into a common column and delete the previous two
```{r}
analysis_clean <- Trauma_data %>%
  mutate(
    nombre = coalesce(`nombre.x`, `nombre.y`),
    nhc = coalesce(`nhc.x`, `nhc.y`),
    vsg_mm_h = coalesce(`vsg_mm_h.x`, `vsg_mm_h.y`),
    crp_serum_mg_d_l = coalesce(`crp_serum_mg_d_l.x`, `crp_serum_mg_d_l.y`),
    d_dimero_plasma_mcg_l_feu = coalesce(
      `d_dimero_plasma_mcg_l_feu.x`,
      `d_dimero_plasma_mcg_l_feu.y`
    )
  ) %>%
  select(
    -`nombre.x`, -`nombre.y`,
    -`nhc.x`, -`nhc.y`,
    -`vsg_mm_h.x`, -`vsg_mm_h.y`,
    -`crp_serum_mg_d_l.x`, -`crp_serum_mg_d_l.y`,
    -`d_dimero_plasma_mcg_l_feu.x`, -`d_dimero_plasma_mcg_l_feu.y`
  )
```

check how the data set changed
```{r}
dim(analysis_clean)
colnames(analysis_clean)
glimpse(analysis_clean)
```

There are lots of useles columns missing values in all the patients, these columns provide no information so we are proceding to delete them.
the empty ones are:

peso
altura
recuento_celular
f_rmula_leucocitaria
cultivo_de_liquido_sinovial
niveles_de_alfa_defensina
resultado_del_analisis_de_corte_congelado
resultados_de_cultivos_de_tejido
confirmacion_microbiologica
microorganismo_identificado

```{r}
missing_overview <- analysis_clean %>%
  summarise(across(
    everything(),
    ~ sum(is.na(.)),
    .names = "{.col}"
  )) %>%
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "n_missing"
  ) %>%
  mutate(
    pct_missing = 100 * n_missing / nrow(analysis_clean)
  ) %>%
  arrange(desc(pct_missing))

missing_overview
```

```{r}
all_missing_cols <- missing_overview %>%
  filter(pct_missing == 100) %>%
  pull(variable)

all_missing_cols
```

```{r}
analysis_qc <- analysis_clean %>%
  select(-all_of(all_missing_cols))
```

Now lets check whether metacluster frequencies sum to 100%, but first see wich columns are from mc frequencies
```{r}
mc_frequency_cols <- setdiff(
  colnames(excel_frequencies),
  c("codigo_interno", "cluster")
)

mc_frequency_cols

length(mc_frequency_cols)

all(mc_frequency_cols %in% colnames(analysis_qc))

setdiff(mc_frequency_cols, colnames(analysis_qc))
```

lets check if the metacluster frequencies sum up to 100%
```{r}
analysis_qc <- analysis_qc %>%
  mutate(
    mc_frequency_sum = rowSums(across(all_of(mc_frequency_cols)), na.rm = TRUE),
    mc_frequency_sum_deviation = mc_frequency_sum - 100
  )

analysis_qc %>%
  select(codigo_interno, mc_frequency_sum, mc_frequency_sum_deviation) %>%
  arrange(desc(abs(mc_frequency_sum_deviation)))
```
Plot the results for a better visualitzation
```{r}
analysis_qc %>%
  ggplot(aes(
    x = reorder(codigo_interno, mc_frequency_sum),
    y = mc_frequency_sum
  )) +
  geom_col() +
  geom_hline(yintercept = 100, linetype = "dashed") +
  coord_flip() +
  labs(
    title = "QC: Sum of FlowSOM metacluster frequencies per patient",
    x = "Patient",
    y = "Sum of metacluster frequencies (%)"
  )
```

After removing useless all-missing variables, what missingness remains in the actual working analysis table?
```{r}
missing_overview_qc <- analysis_qc %>%
  summarise(across(
    everything(),
    ~ sum(is.na(.)),
    .names = "{.col}"
  )) %>%
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "n_missing"
  ) %>%
  mutate(
    pct_missing = round(100 * n_missing / nrow(analysis_qc), 2)
  ) %>%
  arrange(desc(pct_missing), variable)

missing_overview_qc
```
From this info we can extract that "d_dimero_plasma_mcg_l_feu" and  "hb_glicosilada_hb_a1c_percent" variables are not suitable for formal association testing due to high missingness. We will retain "detalle_intervenciones" just as a descriptive field. And "vitamina_d_25_hidroxi_ng_m_l" will be kept but only used as exploratory data. It can be use to describe, maybe plot, but any association with immune populations would be very weak.

The variables found in the 33.33% (10 patients only) missingness can be used but models will have small sample sizes. It would be rather preferable to only use them as descriptive summaries, scatterplots/boxplots, FDR corrections, and to not extract any strong claims from them.

Finally the rest of the varariables that less than 33.33 missignes are the better candidates for association testing.

Creating a table to take decisions on variable usage
```{r}
variable_qc_decisions <- missing_overview_qc %>%
  mutate(
    recommended_use = case_when(
      pct_missing == 100 ~ "Exclude: no observed data",
      pct_missing > 50 ~ "Descriptive only / exclude from formal models",
      pct_missing > 20 & pct_missing <= 50 ~ "Exploratory analysis only",
      pct_missing > 0 & pct_missing <= 20 ~ "Usable with caution",
      pct_missing == 0 ~ "Usable"
    )
  ) %>%
  arrange(desc(pct_missing), variable)

variable_qc_decisions %>%
  print(n = Inf)
```


```{r}
missing_overview_qc %>%
  arrange(desc(pct_missing), variable) %>%
  print(n = Inf)
```

Missingness was concentrated in selected clinical and biochemical variables. All remaining variables, including the FlowSOM metacluster frequency variables, had 0% missingness in the final analysis dataset. Given the small cohort size (n = 15), missingness was interpreted cautiously because each missing observation corresponds to 6.67% of the cohort. And mainly as exploratory.

Before proceding we can obvserve that variables like imc have 0% missignes but only contain "#DIV/0!", which is an excel error message and is hterefore all NA. The same hapens with "ac_ig_m_veb_ea", since all its values are "No procede" it has no analytical variation.
In addition to this "nombre", "nhc", "fecha_de_nacimiento", "descripcion" y "cultivos" are identifiers or text heavy clinical descriptions. These are useful for traceability or manual annotation but not useful for statistical tests.

The following table will store what roles each variable can better perform
```{r}
variable_roles <- tibble(
  variable = colnames(analysis_qc),
  role = case_when(
    variable %in% c("codigo_interno", "nombre", "nhc") ~ "Identifier / traceability only",
    variable %in% c("fecha_de_nacimiento") ~ "Date / traceability only",
    variable %in% c("descripcion", "cultivos", "detalle_intervenciones") ~ "Free-text clinical annotation",
    variable %in% mc_frequency_cols ~ "FlowSOM metacluster frequency",
    variable %in% c("cluster") ~ "FlowSOM-derived patient cluster/group",
    variable %in% c("mc_frequency_sum", "mc_frequency_sum_deviation") ~ "QC-derived variable",
    variable %in% c(
      "edad",
      "numero_intervenciones_n",
      "hemoglobina_g_dl",
      "hematocrito_percent",
      "vcm_f_l",
      "leucocitos_ml_mmc",
      "neutrofilos_percent",
      "linfocitos_percent",
      "vsg_mm_h",
      "crp_serum_mg_d_l",
      "d_dimero_plasma_mcg_l_feu",
      "filtrado_glomerular_m_l_min",
      "proteinas_totales_g_d_l",
      "albumina_g_d_l",
      "prealbumina_mg_d_l",
      "hb_glicosilada_hb_a1c_percent",
      "vitamina_d_25_hidroxi_ng_m_l",
      "il5_pg_m_l",
      "il6_pg_m_l",
      "il10_pg_m_l"
    ) ~ "Numeric clinical/biochemical variable",
    TRUE ~ "Categorical clinical variable"
  )
)

variable_roles %>%
  arrange(role, variable) %>%
  print(n = Inf)
```


```{r}
low_variation_vars <- analysis_qc %>%
  summarise(across(
    everything(),
    ~ n_distinct(., na.rm = TRUE),
    .names = "{.col}"
  )) %>%
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "n_unique_non_missing"
  ) %>%
  arrange(n_unique_non_missing, variable)

low_variation_vars %>%
  print(n = Inf)
```

fixing "imc"
```{r}
analysis_qc <- analysis_qc %>%
  mutate(
    imc_original = imc,
    imc_numeric = suppressWarnings(as.numeric(imc)),
    imc_numeric = if_else(is.nan(imc_numeric), NA_real_, imc_numeric)
  )
```

```{r}
analysis_qc %>%
  select(codigo_interno, imc_original, imc_numeric)
```

```{r}
# Identify columns that are 100% missing
all_na_cols <- analysis_qc %>%
  summarise(across(everything(), ~ all(is.na(.)))) %>%
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "all_na"
  ) %>%
  filter(all_na) %>%
  pull(variable)

all_na_cols
```

```{r}
analysis_qc <- analysis_qc %>%
  select(-all_of(all_na_cols))
```

```{r}
analysis_qc %>%
  summarise(across(everything(), ~ all(is.na(.)))) %>%
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "all_na"
  ) %>%
  filter(all_na)
```

```{r}
low_variation_vars %>%
  print(n = Inf)

variable_roles %>%
  arrange(role, variable) %>%
  print(n = Inf)

```
With the quality cotrol done I proceed to save the outputs
```{r}
saveRDS(
  analysis_qc,
  here("Data", "processed", "atlas_inputs", "analysis_qc_identifiable.rds")
)

write.csv(
  analysis_qc,
  here("Data", "processed", "atlas_inputs", "analysis_qc_identifiable.csv"),
  row.names = FALSE
)
```

```{r}
privacy_cols <- intersect(
  c("nombre", "nhc", "fecha_de_nacimiento", "descripcion"),
  colnames(analysis_qc)
)

analysis_qc_deidentified <- analysis_qc %>%
  select(-all_of(privacy_cols))

saveRDS(
  analysis_qc_deidentified,
  here("Data", "processed", "atlas_inputs", "analysis_qc_deidentified.rds")
)

write.csv(
  analysis_qc_deidentified,
  here("Data", "processed", "atlas_inputs", "analysis_qc_deidentified.csv"),
  row.names = FALSE
)

```

```{r}
saveRDS(
  MC_markers,
  here("Data", "processed", "atlas_inputs", "mc_marker_profiles.rds")
)

write.csv(
  MC_markers,
  here("Data", "processed", "atlas_inputs", "mc_marker_profiles.csv"),
  row.names = FALSE
)
```

```{r}
saveRDS(
  mc_frequency_cols,
  here("Data", "processed", "atlas_inputs", "mc_frequency_cols.rds")
)

writeLines(
  mc_frequency_cols,
  here("Data", "processed", "atlas_inputs", "mc_frequency_cols.txt")
)
```

```{r}
saveRDS(
  mc_frequency_cols,
  here("Data", "processed", "atlas_inputs", "mc_frequency_cols.rds")
)

writeLines(
  mc_frequency_cols,
  here("Data", "processed", "atlas_inputs", "mc_frequency_cols.txt")
)
```

```{r}
mc_frequency_matrix <- analysis_qc_deidentified %>%
  select(codigo_interno, all_of(mc_frequency_cols))

saveRDS(
  mc_frequency_matrix,
  here("Data", "processed", "atlas_inputs", "mc_frequency_matrix.rds")
)

write.csv(
  mc_frequency_matrix,
  here("Data", "processed", "atlas_inputs", "mc_frequency_matrix.csv"),
  row.names = FALSE
)
```

```{r}
patient_metadata <- analysis_qc_deidentified %>%
  select(
    codigo_interno,
    edad,
    sexo,
    lugar_de_infeccion,
    lateralidad,
    tipo_de_infeccion,
    numero_intervenciones_n,
    cluster,
    vsg_mm_h,
    crp_serum_mg_d_l,
    hemoglobina_g_dl,
    leucocitos_ml_mmc,
    neutrofilos_percent,
    linfocitos_percent,
    il5_pg_m_l,
    il6_pg_m_l,
    il10_pg_m_l,
    albumina_g_d_l,
    prealbumina_mg_d_l
  )

saveRDS(
  patient_metadata,
  here("Data", "processed", "atlas_inputs", "patient_metadata.rds")
)

write.csv(
  patient_metadata,
  here("Data", "processed", "atlas_inputs", "patient_metadata.csv"),
  row.names = FALSE
)
```

```{r}
saveRDS(
  missing_overview_qc,
  here("Data", "processed", "atlas_inputs", "missing_overview_qc.rds")
)

saveRDS(
  low_variation_vars,
  here("Data", "processed", "atlas_inputs", "low_variation_vars.rds")
)

saveRDS(
  variable_roles,
  here("Data", "processed", "atlas_inputs", "variable_roles.rds")
)

write.csv(
  missing_overview_qc,
  here("Data", "processed", "atlas_inputs", "missing_overview_qc.csv"),
  row.names = FALSE
)

write.csv(
  low_variation_vars,
  here("Data", "processed", "atlas_inputs", "low_variation_vars.csv"),
  row.names = FALSE
)

write.csv(
  variable_roles,
  here("Data", "processed", "atlas_inputs", "variable_roles.csv"),
  row.names = FALSE
)
```

```{r}
saveRDS(
  variable_qc_decisions,
  here("Data", "processed", "atlas_inputs", "variable_qc_decisions.rds")
)

write.csv(
  variable_qc_decisions,
  here("Data", "processed", "atlas_inputs", "variable_qc_decisions.csv"),
  row.names = FALSE
)
```

```{r}
atlas_input_bundle <- list(
  analysis_qc_deidentified = analysis_qc_deidentified,
  MC_markers = MC_markers,
  mc_frequency_cols = mc_frequency_cols,
  mc_frequency_matrix = mc_frequency_matrix,
  patient_metadata = patient_metadata,
  missing_overview_qc = missing_overview_qc,
  low_variation_vars = low_variation_vars,
  variable_roles = variable_roles,
  session_info = sessionInfo()
)

saveRDS(
  atlas_input_bundle,
  here("Data", "processed", "atlas_inputs", "trauma_immune_atlas_input_bundle.rds")
)
```

