---
title: 'Creation of FlowSOM immune atlas: Prosthetic Infection Immune Profiling'
author: "Bruno Álvarez"
date: "2026-05-14"
output:
  pdf_document: 
    toc: true
    number_sections: true
    latex_engine: xelatex
  geometry: margin=1n
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)

library(tidyverse)
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(janitor)
library(ggplot2)
library(skimr)
library(naniar)
library(knitr)
library(kableExtra)
library(here)
library(writexl)
library(pheatmap)
library(ggrepel)
library(broom)
library(patchwork)
library(uwot)
library(tidyverse)
```

The goal of the immune atlas is to properly describe the immune landscape reported by FlowSOM meta clusters, biologically annotated those meta clusters, and visualize how the 15 patients distribute across those different immune states.
The main objectives that this report wants to achieve are the following ones:
- Portray what immune populations did FlowSOM identify
- Identify what marker patterns define each meta cluster
- How metaclusters look like at a cellular level
- How abundant is each population across the different patients
- And answer if patients show broad immune-profile structure


First step, loading the data saved in the QC section
```{r}
bundle <- readRDS("Data/processed/atlas_inputs/trauma_immune_atlas_input_bundle.rds")
```

```{r}
names(bundle)
```

```{r}
#extract objects
analysis_qc_deidentified <- bundle$analysis_qc_deidentified
mc_marker_profiles       <- bundle$MC_markers
mc_frequency_cols        <- bundle$mc_frequency_cols
mc_frequency_matrix      <- bundle$mc_frequency_matrix
patient_metadata         <- bundle$patient_metadata
missing_overview_qc      <- bundle$missing_overview_qc
low_variation_vars       <- bundle$low_variation_vars
variable_roles           <- bundle$variable_roles
```

```{r}
dim(analysis_qc_deidentified)
dim(mc_marker_profiles)
length(mc_frequency_cols)
dim(mc_frequency_matrix)
dim(patient_metadata)
```

We proceed to creat a heatmap for the for the 20 meta clusters in respect to the patients.

First creating an specific table to plot the heatmap
```{r}

patient_immune_tbl <- analysis_qc_deidentified %>%
  as_tibble() %>%
  mutate(
    patient_row_id = row_number(),
    patient_plot_id = paste0("P", stringr::str_pad(patient_row_id, width = 2, pad = "0"))
  ) %>%
  select(
    codigo_interno,
    patient_row_id,
    patient_plot_id,
    cluster,
    all_of(mc_frequency_cols)
  )

patient_immune_tbl
```

Transforming the immune table into long format
```{r}
patient_mc_long <- patient_immune_tbl %>%
  pivot_longer(
    cols = all_of(mc_frequency_cols),
    names_to = "metacluster",
    values_to = "frequency"
  ) %>%
  mutate(
    cluster = as.factor(cluster)
  )

head(patient_mc_long)
```
Ploting the raw frequencie heatmap (Patient x FlowSOM metacluster frequency heatmap), patients being frouped by the FlowSOM derived patient cluster.

```{r, fig.crop=FALSE}
ggplot(patient_mc_long, aes(x = metacluster, y = codigo_interno, fill = frequency)) +
  geom_tile(color = "white", linewidth = 0.2) +
  facet_grid(cluster ~ ., scales = "free_y", space = "free_y") +
  labs(
    x = "Metacluster",
    y = "Patient",
    fill = "Frequency (%)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    plot.margin = margin(t = 25, r = 0, b = 0, l = 0)
    
  )
```

In the raw patient × metacluster frequency heatmap, most metaclusters show very low frequencies across patients. Cluster 1 appears characterized by higher abundance of cd8_tem_ra_cd57dim_senescent. Cluster 2 shows an attenuated abundance of this metacluster and relatively higher abundance of cd4_naive_ccr7hi_cd27_cd28. Cluster 3 contains only one patient and appears as a single-patient outlier profile, dominated by cd4dim_pd1hi_tim3hi_exhausted.

In order to not rely entirely on the heatmap now we are going to quantify metacluster frequencies by FlowSOM patient cluster
```{r}
cluster_mc_summary <- patient_mc_long %>%
  group_by(cluster, metacluster) %>%
  summarise(
    n_patients = n(),
    mean_frequency = mean(frequency, na.rm = TRUE),
    median_frequency = median(frequency, na.rm = TRUE),
    sd_frequency = sd(frequency, na.rm = TRUE),
    min_frequency = min(frequency, na.rm = TRUE),
    max_frequency = max(frequency, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(cluster, desc(mean_frequency))

cluster_mc_summary
```

Now we create a dataset focused only on the metaclusters of interes or the ones that may have a high frequency and we didnt notice
```{r}
mcs_of_interest <- c(
  "cd8_tem_ra_cd57dim_senescent",
  "cd4_quiescent_all_markers_low",
  "cd4dim_pd1hi_tim3hi_exhausted"
)

#checking that these names exist exactly in the data
setdiff(mcs_of_interest, unique(patient_mc_long$metacluster))

focused_mc_summary <- patient_mc_long %>%
  filter(metacluster %in% mcs_of_interest) %>%
  group_by(cluster, metacluster) %>%
  summarise(
    n_patients = n(),
    mean_frequency = mean(frequency, na.rm = TRUE),
    median_frequency = median(frequency, na.rm = TRUE),
    min_frequency = min(frequency, na.rm = TRUE),
    max_frequency = max(frequency, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(metacluster, cluster)

focused_mc_summary
```

```{r}
ggplot(
  focused_mc_summary,
  aes(x = metacluster, y = mean_frequency, fill = as.factor(cluster))
) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(
    x = "Metacluster",
    y = "Mean frequency (%)",
    fill = "Patient cluster"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

From this new information we can say that Cluster 3 shows high frequency/aubndance of the cd4dim_pd1hi_tim3hi_exhausted 
Then for cluster 1 we see a dominance by cd8_tem_ra_cd57dim_senescent (54.84%), suggesting a strong CD8 TEMRA/senescent-like skewed immune profile.
On the other hand, cluster 2 looks dominated by cd4_quiescent_all_markers_low (35.65%), with an intermediate contribution from cd8_tem_ra_cd57dim_senescent (25.87%).
Cluster 3 contains only one patient and is dominated by cd4dim_pd1hi_tim3hi_exhausted, suggesting a single-patient outlier immune profile enriched in a PD1/TIM-3-high CD4dim metacluster.
cd4_naive_ccr7hi_cd27_cd28 was present in both clusters and showed only a modest difference between clusters 1 and 2 of 10.56% in cluster 2 against 7.17% cluster 1.

The focused bar plot confirmed that the three FlowSOM-derived patient clusters were mainly separated by three dominant metacluster-frequency patterns.

For `cd4_quiescent_all_markers_low`, the mean frequency was highest in cluster 2, reaching almost 40%, intermediate in cluster 1 at around 20%, and almost absent in cluster 3. This suggests that cluster 2 is the patient group most enriched in this CD4 quiescent-like metacluster.

For `cd4dim_pd1hi_tim3hi_exhausted`, the frequency was approximately 0% in clust ers 1 and 2, while cluster 3 showed a high frequency of approximately 75%. Since cluster 3 contains only one patient, this pattern should be interpreted as a single-patient outlier immune profile dominated by a PD1/TIM-3-high CD4dim metacluster, rather than as a robust group-level phenotype.

For `cd8_tem_ra_cd57dim_senescent`, the mean frequency was highest in cluster 1, close to 60%, intermediate in cluster 2, around 30%, and nearly absent in cluster 3. This supports the interpretation that cluster 1 is characterized by a CD8 TEMRA/senescent-like dominant immune profile, while cluster 2 retains a moderate contribution from this metacluster.

Overall, the focused bar plot suggests a gradient-like distinction between clusters 1 and 2, with cluster 1 being more CD8 TEMRA/senescent-like and cluster 2 being more CD4 quiescent-like. Cluster 3 differs qualitatively from both clusters because it is dominated by the `cd4dim_pd1hi_tim3hi_exhausted` metacluster and contains only one patient.

Now we can continue and perform PCAs, we are gona use them on the FlowSOM MC frequencies to see if patients separate based on their immue metacluster frequency profiles.
Afterwards we will color them based on the FlowSOM clusters and on the clinical data.

However, before performing the PCAs there is an issue with the data. THe frequencies in the meta clusters sum up to 100 for each patient, so if one meta cluster goes up at least one other must go down mathematically, eventhough if biologically nothing has decreased in an absolute sense. This type of percentages are called compositional.
This causes to think between patients for example in this case:

Patient A:
MC1 = 60%, MC2 = 40%

Patient B:
MC1 = 80%, MC2 = 20%

that MC2 decrased, however it may be that only MC1 expanded and MC2 absolut count stayed similar.With percentages alone we can only know relative composition and not absolute cell count.
So for a PCA distances, clustering and ordination should be transformed before using. I will therefore tranform them through CLR trasnformation (centered log-ratio transformation).
That is because CLR helps deal with the fact that the data are proportions, making the PCA more appropriate.

However CLR has an issue, due to CLR using logarithms it can not handle zeros. That is why we have to add a tiny pseudocount before transforming our data.

CLR TRANSFORMATION:

we separate frequencies to another separate matrix
```{r}
mc_frequency_raw <- patient_immune_tbl %>%
  select(all_of(mc_frequency_cols))
```

we check for 0 values
```{r}
zero_check <- mc_frequency_raw %>%
  summarise(
    total_values = n() * ncol(.),
    zero_values = sum(. == 0, na.rm = TRUE),
    missing_values = sum(is.na(.))
  )

zero_check
```
We transform percentages to proportions, by instead of simply dividing by 100, dividing by each patient's total frequency. This is a much safer aproach because rows sum to approximately 100 but not always 100. And force each patient row to sum to 1.
```{r}
mc_frequency_prop <- mc_frequency_raw / rowSums(mc_frequency_raw, na.rm = TRUE)

#checking row sums
summary(rowSums(mc_frequency_prop, na.rm = TRUE))
```
All of them should sum 1 so min max and median should be 1.

Due to CLR not being able to take 0 due to log(0) being undefined.
A reasonable choice is to add half of the smallest non zero proportion.
```{r}
min_nonzero_prop <- min(as.matrix(mc_frequency_prop)[as.matrix(mc_frequency_prop) > 0], na.rm = TRUE)

pseudocount <- min_nonzero_prop / 2

pseudocount
```

Now we add the pseudocount and normalize the rows so that they sum up to 1 again
```{r}
mc_frequency_prop_pseudo <- mc_frequency_prop + pseudocount

mc_frequency_prop_pseudo <- mc_frequency_prop_pseudo / rowSums(
  mc_frequency_prop_pseudo,
  na.rm = TRUE
)


#checking the wows aggain after pseudocount
summary(rowSums(mc_frequency_prop_pseudo, na.rm = TRUE))
```

Now we can apply the CLR transformation. the CLR formula that we are going to apply is the following one:
CLR(x_i) = log(x_i) - mean(log(x_1), log(x_2), ..., log(x_D))
```{r}
#CLR transform function
clr_transform <- function(x) {
  log_x <- log(x)
  log_x - mean(log_x, na.rm = TRUE)
}

```

```{r}
#applying formula
mc_frequency_clr <- mc_frequency_prop_pseudo %>%
  as.matrix() %>%
  t() %>%
  apply(2, clr_transform) %>%
  t() %>%
  as.data.frame()

#restoring the names of the columns
colnames(mc_frequency_clr) <- mc_frequency_cols

#inspecting result
dim(mc_frequency_clr)
```

Verifying the result
```{r}
clr_row_sum_check <- rowSums(mc_frequency_clr)

summary(clr_row_sum_check)
```

Now we can go back to having a patient table but with the metacluster values transformed
```{r}
patient_immune_clr_tbl <- patient_immune_tbl %>%
  select(patient_row_id, patient_plot_id, cluster, codigo_interno) %>%
  bind_cols(mc_frequency_clr)

patient_immune_clr_tbl
```

Now we can do the PCA
```{r}
#creating the input for the PCA algorithm
pca_input_matrix <- patient_immune_clr_tbl %>%
  select(all_of(mc_frequency_cols)) %>%
  as.matrix()
```

```{r}
#Running the PCA
pca_res <-prcomp(
  pca_input_matrix,
  center = TRUE,
  scale. = FALSE
)
```

Check the variance explained
```{r}
pca_variance <- tibble(
  PC = paste0("PC", seq_along(pca_res$sdev)),
  variance_explained = (pca_res$sdev^2) / sum(pca_res$sdev^2),
  percent_variance = variance_explained * 100
)
```

now we generate the PCA scores
```{r}
pca_scores <- as_tibble(pca_res$x) %>%
  bind_cols(
    patient_immune_clr_tbl %>%
      select(patient_row_id, patient_plot_id, cluster, codigo_interno)
  ) %>%
  mutate(cluster = as.factor(cluster))

pca_scores
```

Now we can finally plot our PCA and we will colour it using the already existing FlowSOM patient clusters
```{r}
ggplot(pca_scores, aes(x = PC1, y = PC2, color = cluster, label = codigo_interno)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.8, size = 3, show.legend = FALSE) +
  labs(
    x = paste0("PC1 (", round(pca_variance$percent_variance[1], 1), "%)"),
    y = paste0("PC2 (", round(pca_variance$percent_variance[2], 1), "%)"),
    color = "Patient cluster"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank()
  )
```

PCA of CLR-transformed metacluster frequencies did not show a strong separation of the three FlowSOM-derived patient clusters, suggesting that the cluster structure is not fully captured by the first two principal components.

PC1 and PC2 have enough variance_explained (3.368679e-01 and 2.589606e-01), however separation is sometimes not visible comparing between PC1 and PC2 so I'm going to try adding PC3 to the mixt.
```{r}
ggplot(pca_scores, aes(x = PC1, y = PC3, color = cluster, label = codigo_interno)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.8, size = 3, show.legend = FALSE) +
  labs(
    x = paste0("PC1 (", round(pca_variance$percent_variance[1], 1), "%)"),
    y = paste0("PC3 (", round(pca_variance$percent_variance[3], 1), "%)"),
    color = "Patient cluster"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank()
  )
```
We can see a 'phantom' tendency of cluster 2 tending to the negative values in the x and y scale and and for cluster 1 to do the opposite, however this is not a conclusion or causal analysis its only an exploratory analysis and subjective observation.
However in a global analysis clusters are mixed and PCA does not strongly support global separation.

Lets further inspect this finding by identifying which metaclusters drive PC1, PC2, and PC3.
```{r}
# PCA loadings
pca_loadings <- as_tibble(
  pca_res$rotation,
  rownames = "metacluster"
)

pca_loadings %>%
  select(metacluster, PC1, PC2, PC3) %>%
  arrange(desc(abs(PC1)))
```

the top ones we select them in absolute values since PCA joins the meta clusters that affect the most moving the value positively and negatively, so we want the top of both ways to see which ones affect the most
```{r}
top_pc1_loadings <- pca_loadings %>%
  select(metacluster, PC1) %>%
  mutate(abs_loading = abs(PC1)) %>%
  arrange(desc(abs_loading)) %>%
  slice_head(n = 10)

top_pc1_loadings
```

```{r}
top_pc2_loadings <- pca_loadings %>%
  select(metacluster, PC2) %>%
  mutate(abs_loading = abs(PC2)) %>%
  arrange(desc(abs_loading)) %>%
  slice_head(n = 10)

top_pc2_loadings
```
lets plot the top drivers for PC1 and PC2
```{r}
top_loadings <- pca_loadings %>%
  select(metacluster, PC1, PC2) %>%
  pivot_longer(
    cols = c(PC1, PC2),
    names_to = "PC",
    values_to = "loading"
  ) %>%
  group_by(PC) %>%
  slice_max(order_by = abs(loading), n = 10) %>%
  ungroup()

ggplot(top_loadings, aes(x = reorder(metacluster, loading), y = loading)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ PC, scales = "free_y") +
  labs(
    x = "Metacluster",
    y = "PCA loading"
  ) +
  theme_minimal(base_size = 11)
```

This table was used to interpret the PCA loadings. Positive loadings indicate metaclusters associated with patients positioned toward positive values of a principal component, whereas negative loadings indicate metaclusters associated with patients positioned toward negative values. Therefore, each principal component can be interpreted as a contrast between the metaclusters contributing most strongly in opposite directions.

So we can see between which two profiles PC1 and PC2 separates patients.
As we can see PC1 is mainly driven mainly by gd_t_tem_ra_cd45r_ahi_vd1 in the positive direction and cd4_naive_ccr7hi_cd27_cd28 in the negative.
This suggests that PC1 captures a contrast between a γδ T-cell TEMRA-like Vδ1-associated profile and a CD4 naïve-like profile.


On the other hand, for PC2 the two main meta clusters influencing it are cd4dim_pd1hi_tim3hi_exhausted to the positive side, this one remarkably contributes compared to the rest and to the PC1 by far more. For the negative gd_t_mixed_vd_vd2dim.
The contribution of cd4dim_pd1hi_tim3hi_exhausted was particularly strong, consistent with the outlier profile observed in patient cluster 3.

The patient × metacluster heatmap and focused bar plots showed clear differences in selected dominant metaclusters between the FlowSOM-derived patient clusters. In particular, cluster 1 was characterized by higher `cd8_tem_ra_cd57dim_senescent`, cluster 2 by higher `cd4_quiescent_all_markers_low`, and cluster 3 by high `cd4dim_pd1hi_tim3hi_exhausted`.

However, the PCA showed a more mixed and dispersed patient distribution. This indicates that although specific metaclusters differ between patient clusters, the global 20-metacluster immune profiles do not separate into completely distinct groups along PC1 and PC2. Therefore, the FlowSOM patient clusters should be interpreted as exploratory immune-profile patterns rather than sharply separated or validated immunotypes.

This difference is expected because the heatmap/bar plots focus on selected dominant metaclusters, whereas PCA summarizes variation across the full 20-metacluster composition.

Due to this dispersion I will then now proceed to run a sanity check. We are going to perform hierarchical clustering. Which is an unsupervised patient grouping method. This method creates a dendogram showcasing which patients have a similar immune profile in global.
The main goal of this is not to replace the already existing "Cluster" variable, but to check whether an independent clustering method gives patient groupings similar to the FlowSOM ones.

If it matches it would help support the FlowSOM cluster structure. However, if it did not match it would mean that FlowSOM clusters capture selected dominant meta cluster differences, but the global profile of the 20 meta clusters is more mixed. 
This analysis is exploratory and was not used to redefine patient groups.

Lets first prepare the clustering input
```{r}
hc_input_matrix <- patient_immune_clr_tbl %>%
  select(all_of(mc_frequency_cols)) %>%
  as.matrix()

rownames(hc_input_matrix) <- patient_immune_clr_tbl$codigo_interno

dim(hc_input_matrix)
```

Now we can compute the distances and the clustering
```{r}
#Euclidean distance on CLR-transformed immune profiles
patient_dist <- dist(hc_input_matrix, method = "euclidean")

#hierarchical clustering
hc_res <- hclust(patient_dist, method = "ward.D2")
```

Ploting the dendrogram
```{r}
plot(
  hc_res,
  main = "Hierarchical clustering of CLR immune profiles",
  xlab = "Patient",
  sub = "",
  cex = 0.8
)

rect.hclust(hc_res, k = 3, border = "gray40")
```


Comparing the hierarchical clusters with the existing FlowSOM clusters
```{r}
hc_patient_groups <- tibble(
  codigo_interno = names(cutree(hc_res, k = 3)),
  hc_cluster = as.factor(cutree(hc_res, k = 3))
) %>%
  left_join(
    patient_immune_clr_tbl %>%
      select(codigo_interno, flowsom_patient_cluster = cluster),
    by = "codigo_interno"
  ) %>%
  mutate(
    flowsom_patient_cluster = as.factor(flowsom_patient_cluster)
  )

hc_patient_groups
```
Now we visualize with a comparison table or with a matrix
```{r}
hc_comparison_table <- hc_patient_groups %>%
  count(flowsom_patient_cluster, hc_cluster)

hc_comparison_table
```

```{r}
table(
  FlowSOM_cluster = hc_patient_groups$flowsom_patient_cluster,
  Hierarchical_cluster = hc_patient_groups$hc_cluster
)
```

So we can observe that the hierarchical clustering fitted every patient onto cluster 1 and left 2 patients out into one cluster each one. consequently we can conclude that hierarchical clustering did not fully reproduce the FlowSOM-derived patient clusters, suggesting that the patient clusters capture selected metacluster-dominant patterns rather than sharply separated global immune-profile groups.

To continue with our descriptive analysis we now can use clinical metadata overlays on the PCA to paint the samples and see if theres any information worth distinguishing.

Which clinical variables do we have available?
```{r}
names(patient_metadata)
```
We now check the missingness to choose the best variables to test
```{r}
patient_metadata %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "n_missing"
  ) %>%
  arrange(n_missing)
```
We now join the clinical data to the pca scores
```{r}
pca_scores_metadata <- pca_scores %>%
  left_join(
    patient_metadata %>% select(-cluster),
    by = "codigo_interno"
  )
```

Checking the fusion
```{r}
anti_join(pca_scores, patient_metadata, by = "codigo_interno")
anti_join(patient_metadata, pca_scores, by = "codigo_interno")
```
Now we can do an exploratory analysis on how immune 
```{r}
pca_scores_metadata_plot <- pca_scores_metadata %>%
  mutate(
    sexo = as.factor(sexo),
    lugar_de_infeccion = as.factor(lugar_de_infeccion),
    tipo_de_infeccion = as.factor(tipo_de_infeccion),
    numero_intervenciones_n = as.factor(numero_intervenciones_n),
    lateralidad = as.factor(lateralidad),
    il6_pg_m_l = il6_pg_m_l
  )

p_sexo <- ggplot(pca_scores_metadata_plot, aes(x = PC1, y = PC2, color = sexo, label = codigo_interno)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.8, size = 3, show.legend = FALSE) +
  labs(
    x = paste0("PC1 (", round(pca_variance$percent_variance[1], 1), "%)"),
    y = paste0("PC2 (", round(pca_variance$percent_variance[2], 1), "%)"),
    color = "sex"
  ) +
  theme_minimal(base_size = 11)


p_lugar <- ggplot(pca_scores_metadata_plot, aes(x = PC1, y = PC2, color = lugar_de_infeccion, label = codigo_interno)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.8, size = 3, show.legend = FALSE) +
  labs(
    x = paste0("PC1 (", round(pca_variance$percent_variance[1], 1), "%)"),
    y = paste0("PC2 (", round(pca_variance$percent_variance[2], 1), "%)"),
    color = "place of infection"
  ) +
  theme_minimal(base_size = 11)

p_tipo <- ggplot(pca_scores_metadata_plot, aes(x = PC1, y = PC2, color = tipo_de_infeccion, label = codigo_interno)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.8, size = 3, show.legend = FALSE) +
  labs(
    x = paste0("PC1 (", round(pca_variance$percent_variance[1], 1), "%)"),
    y = paste0("PC2 (", round(pca_variance$percent_variance[2], 1), "%)"),
    color = "type of infection"
  ) +
  theme_minimal(base_size = 11)

p_intervenciones <- ggplot(pca_scores_metadata_plot, aes(x = PC1, y = PC2, color = numero_intervenciones_n, label = codigo_interno)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.8, size = 3, show.legend = FALSE) +
  labs(
    x = paste0("PC1 (", round(pca_variance$percent_variance[1], 1), "%)"),
    y = paste0("PC2 (", round(pca_variance$percent_variance[2], 1), "%)"),
    color = "number of interventions"
  ) +
  theme_minimal(base_size = 11)

```

```{r}
(p_sexo + p_lugar) /
(p_tipo + p_intervenciones)
```
In general no big distinctions can be found, as an example are the four figures above. Despite that there are two really subtle cases in which something could be commented.

```{r}
lateralidad_p <- ggplot(pca_scores_metadata_plot, aes(x = PC1, y = PC2, color = lateralidad, label = codigo_interno)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.8, size = 3, show.legend = FALSE) +
  labs(
    x = paste0("PC1 (", round(pca_variance$percent_variance[1], 1), "%)"),
    y = paste0("PC2 (", round(pca_variance$percent_variance[2], 1), "%)"),
    color = "lateralidad"
  ) +
  theme_minimal(base_size = 11)

il6_p <- ggplot(pca_scores_metadata_plot, aes(x = PC1, y = PC2, color = il6_pg_m_l, label = codigo_interno)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.8, size = 3, show.legend = FALSE) +
  labs(
    x = paste0("PC1 (", round(pca_variance$percent_variance[1], 1), "%)"),
    y = paste0("PC2 (", round(pca_variance$percent_variance[2], 1), "%)"),
    color = "il6_pg_m_l"
  ) +
  theme_minimal(base_size = 11)

(lateralidad_p + il6_p)
```
A weak visual trend was observed in the case of laterality that the patients with "Derecha" (right) tend to the left, the only patient with "ambas" (both) falls around the middle and the ones with "izquierda" (left) fall more to the l¡right in respect to PC1, however this is a really rough extrapolation and should be interpreted cautiously. 

In the case of il6_pg_m_l we can see that low values cluster in the center, while higher ones disperse in all directions.

Now, and finally, to give another point of view and keep on going with our exploratory visualization we can perform a UMAP, an algorithm with the same aim as the PCA but with a different approach. In this case it performs a dimensionality reduction trying to keep the neighbors.


We start by preparing the input to the UMAP
```{r}
umap_input_matrix <- patient_immune_clr_tbl %>%
  select(all_of(mc_frequency_cols)) %>%
  as.matrix()

rownames(umap_input_matrix) <- patient_immune_clr_tbl$codigo_interno

dim(umap_input_matrix)
```

Now we run the UMAP
```{r}
set.seed(123) #setting a seed because contrary to the parametric UMAP, normal UMAP is not replicable

umap_res <- uwot::umap(
  umap_input_matrix,
  n_neighbors = 4, #we have small sample size
  min_dist = 0.3,
  metric = "euclidean", #apropiate for clr transformation
  scale = FALSE #data is already trasnformed with clr
)
```

Creating the UMAP plotting table
```{r}
umap_scores <- as_tibble(umap_res, .name_repair = "minimal") %>%
  rename(
    UMAP1 = 1,
    UMAP2 = 2
  ) %>%
  bind_cols(
    patient_immune_clr_tbl %>%
      select(codigo_interno, cluster)
  ) %>%
  mutate(
    cluster = as.factor(cluster)
  )

umap_scores
```

Plotting the UMAP colored by FlowSOM clusters
```{r}
ggplot(umap_scores, aes(x = UMAP1, y = UMAP2, color = cluster, label = codigo_interno)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.8, size = 3, show.legend = FALSE) +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    color = "Patient cluster"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank()
  )
```
Again we can observe that when the whole immune landscape is used to classify all samples appear mixed

We can join with the clinical metadata to check the "lateralidad" and "il6_pg_m_l" variables like before.
```{r}
umap_scores_metadata <- umap_scores %>%
  left_join(
    patient_metadata %>% select(-cluster),
    by = "codigo_interno"
  )

```

```{r}
ggplot(umap_scores_metadata, aes(x = UMAP1, y = UMAP2, color = il6_pg_m_l)) +
  geom_point(size = 3.5, alpha = 0.9) +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    color = "IL-6\npg/mL"
  ) +
  theme_minimal(base_size = 11)
```
For "il6_pg_m_l" the UMAP overlay did not show a clear or consistent visual pattern

```{r}
ggplot(umap_scores_metadata, aes(x = UMAP1, y = UMAP2, color = lateralidad)) +
  geom_point(size = 3.5, alpha = 0.9) +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    color = "IL-6\npg/mL"
  ) +
  theme_minimal(base_size = 11)
```
In the case of "lateralidad", we can again see a visual separation between "Derecha" (right) and "Izquierda" (left). "Derecha" tending to occupy the lower region of the UMAP spac, and "Izquierda" in the upper one. 
This observation should be read with caution, since there is no immediate biological rationale that would cause to expect immune profile structure to be driven by the laterality of the injure. This could be due to the instability of UMAP in a cohort of only 15 patients however it is courious to see that it also happened in the PCA.



Overall, this immune-atlas analysis describes the patient-level immune landscape captured by FlowSOM metacluster frequencies. The FlowSOM-derived patient clusters were mainly characterized by differences in a limited number of dominant metaclusters rather than by complete separation across the full immune profile. Cluster 1 showed a CD8 TEMRA/senescent-like dominant pattern, mainly driven by `cd8_tem_ra_cd57dim_senescent`; cluster 2 showed a more CD4 quiescent-like mixed profile, mainly associated with `cd4_quiescent_all_markers_low`; and cluster 3 consisted of a single-patient outlier profile dominated by `cd4dim_pd1hi_tim3hi_exhausted`.

After CLR transformation of the compositional metacluster-frequency data, PCA and UMAP showed that patients were globally mixed rather than sharply separated into stable immune groups. Hierarchical clustering also did not fully reproduce the FlowSOM-derived patient clusters, supporting the interpretation that these clusters reflect exploratory immune-profile patterns rather than validated immunotypes. Clinical metadata overlays did not reveal strong or consistent associations with the main immune-profile axes. Weak visual trends, such as the apparent separation by `lateralidad`, were considered hypothesis-generating only and should be interpreted with caution due to the small cohort size and unbalanced categories.

Therefore, this section provides a descriptive FlowSOM-based immune atlas of the cohort. The results identify dominant immune profile patterns and candidate metacluster signatures, but they should not be interpreted as definitive clinical biomarkers, causal mechanisms, or validated patient subtypes. Further analysis with larger cohorts and formal statistical testing would be required to evaluate clinical associations.
The main following steps would be an analysis involving a cohort quality audit, a global map of the patients and perhaps a differential abundance.


Saving the important objects
```{r}
immune_atlas_results <- list(
  patient_immune_tbl = patient_immune_tbl,
  patient_mc_long = patient_mc_long,
  cluster_mc_summary = cluster_mc_summary,
  focused_mc_summary = focused_mc_summary,
  mc_frequency_clr = mc_frequency_clr,
  patient_immune_clr_tbl = patient_immune_clr_tbl,
  pca_res = pca_res,
  pca_scores = pca_scores,
  pca_variance = pca_variance,
  pca_loadings = pca_loadings,
  umap_scores = umap_scores
)

saveRDS(
  immune_atlas_results,
  "trauma_immune_atlas_results_bundle.rds"
)
```